Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-04-13 doc/talks/3_long/talk.tex
First draft of the entire talk

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}
% TODO: unify the use of colours, maybe replace red!50 with one of the main colours
% TODO: how do we capitalise titles of \block?
% TODO: make sure I'm using the 'forest' package whenever possible
% TODO: consider adding an outline slide
% TODO: add periods at the end of sentences

@@

      \begin{block}{Generalisations of the weight function}

@@

\begin{frame}{(Unweighted) First-Order Model Counting}
  \begin{itemize}
    \item Example formula: \\
          \alert{$\forall x \in \Delta\text{. } \pP(x) \lor \qQ(x)$}
    \item Let \alert{$\Delta \coloneqq \{\, 1, 2 \,\}$}
    \item Interpretations: all subsets of\\
          \alert{$\{\, \pP(1), \qQ(1), \pP(2), \qQ(2) \,\}$}
    \item<2-> Models:
    \[
      \begin{matrix}
        \{\, \pone, \ptwo \,\}, & \{\, \pone, \qtwo \,\}, & \{\, \pone, \pqtwo \,\},\\
        \{\, \qone, \ptwo \,\}, & \{\, \qone, \qtwo \,\}, & \{\, \qone, \pqtwo \,\},\\
        \{\, \pqone, \ptwo \,\}, & \{\, \pqone, \qtwo \,\}, & \{\, \pqone, \pqtwo \,\}\\
      \end{matrix}
    \]
  \end{itemize}
  \onslide<3->{
    \begin{block}{Intuition}
      \begin{itemize}
        \item Each 1-ary predicate is like a \alert{subset}
        \item For \alert{$n>1$}, each \alert{$n$}-ary predicate is like a
              \alert{relation}
        \item FOMC counts \alert{combinations of relations}
      \end{itemize}
    \end{block}
  }
  \begin{tikzpicture}[remember picture,overlay,visible on=<2->]
    \tikzset{shift={(current page.north east)}}
    \tikzset{shift={(-4, -2.9)}}
    \node[draw,circle,minimum size=3cm,label={135:$\pP$},fill=color1] (P) at (0,0) {};
    \node[draw,circle,minimum size=3cm,label={45:$\qQ$},fill=color2] (Q) at (1.8,0) {};
    \begin{scope}
      \clip (0,0) circle(1.5cm);
      \clip (1.8,0) circle(1.5cm);
      \fill[color3](0,0) circle(1.5cm);
    \end{scope}
    \draw (0,0) circle(1.5cm);
    \draw (1.8,0) circle(1.5cm);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a \alert{$\#\P$-complete} problem
          \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

@@

  Our new rules enable Crane to efficiently count
  \alert{$\Gamma \to \Delta$} functions such as:
  \begin{itemize}
    \item injections in \alert{$\Theta(mn)$} time
          \begin{itemize}
            \item by hand: \alert{$\Theta(m)$}
          \end{itemize}
    \item partial injections in \alert{$\Theta(mn)$} time
          \begin{itemize}
            \item by hand: \alert{$\Theta({\min\{\, m, n \,\}}^2)$}
          \end{itemize}
          \item bijections in \alert{$\Theta(m)$} time
          \begin{itemize}
            \item \alert{optimal!}
            \item FastWFOMC seems to be \alert{$\Omega(m^{4})$}
          \end{itemize}
  \end{itemize}
\end{frame}
% TODO: add 2 more statements about fastwfomc, showing their plots on the RHS
% when revealing them

@@

% TODO: maybe add a future <-> past picture slide

@@

\begin{frame}{Algebraic Solutions to Parameterised Problems (2/2)}
  With the help of a computer algebra system, we can then:
  \begin{itemize}
    \item determine how \alert{$P$} scales with the number of users,
    \item find combinations of domain sizes that keep \alert{$P$} below some
          threshold,
    \item find ranges of weights that keep \alert{$P$} sufficiently small across
          a range of domain size values.
  \end{itemize}
\end{frame}
% TODO: add a concluding diagram: 'reasoning with functions'(in a box) >(big) 'reasoning with numbers'

>>>>

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}

@@

      \begin{block}{Generalisations of the Weight Function}

@@

\begin{frame}{(Unweighted) First-Order Model Counting}
  \begin{itemize}
    \item Example formula: \\
          \alert{$\forall x \in \Delta\text{. } \pP(x) \lor \qQ(x)$}.
    \item Let \alert{$\Delta \coloneqq \{\, 1, 2 \,\}$}.
    \item Interpretations: all subsets of\\
          \alert{$\{\, \pP(1), \qQ(1), \pP(2), \qQ(2) \,\}$}.
    \item<2-> Models:
    \[
      \begin{matrix}
        \{\, \pone, \ptwo \,\}, & \{\, \pone, \qtwo \,\}, & \{\, \pone, \pqtwo \,\},\\
        \{\, \qone, \ptwo \,\}, & \{\, \qone, \qtwo \,\}, & \{\, \qone, \pqtwo \,\},\\
        \{\, \pqone, \ptwo \,\}, & \{\, \pqone, \qtwo \,\}, & \{\, \pqone, \pqtwo \,\}.\\
      \end{matrix}
    \]
  \end{itemize}
  \onslide<3->{
    \begin{block}{Intuition}
      \begin{itemize}
        \item Each 1-ary predicate is like a \alert{subset}.
        \item For \alert{$n>1$}, each \alert{$n$}-ary predicate is like a
              \alert{relation}.
        \item FOMC counts \alert{combinations of relations}.
      \end{itemize}
    \end{block}
  }
  \begin{tikzpicture}[remember picture,overlay,visible on=<2->]
    \tikzset{shift={(current page.north east)}}
    \tikzset{shift={(-4, -2.9)}}
    \node[draw,circle,minimum size=3cm,label={135:$\pP$},fill=color1] (P) at (0,0) {};
    \node[draw,circle,minimum size=3cm,label={45:$\qQ$},fill=color2] (Q) at (1.8,0) {};
    \begin{scope}
      \clip (0,0) circle(1.5cm);
      \clip (1.8,0) circle(1.5cm);
      \fill[color3](0,0) circle(1.5cm);
    \end{scope}
    \draw (0,0) circle(1.5cm);
    \draw (1.8,0) circle(1.5cm);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming.
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}.
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a \alert{$\#\P$-complete} problem.
          \pause
    \item The rest of this talk is about how to find such functions
          automatically.
  \end{itemize}
\end{frame}

@@

  Our new rules enable Crane to efficiently count \alert{$\Gamma \to \Delta$}
  functions such as:\pause
  \begin{itemize}
    \item injections in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item by hand: \alert{$\Theta(m)$}\pause
    \end{itemize}
    \item partial injections in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item by hand: \alert{$\Theta({\min\{\, m, n \,\}}^2)$}\pause
    \end{itemize}
    \item bijections in \alert{$\Theta(m)$} time
    \begin{itemize}
      \item \alert{optimal!}\pause
      \item experimentally, FastWFOMC runtime appears to be \alert{$\Omega(m^{4})$}
    \end{itemize}
  \end{itemize}
  \begin{tikzpicture}[remember picture,overlay]
    \node[draw,rectangle,color=gray] at ([xshift=-3cm]current page.east) {\includegraphics{plot}};
  \end{tikzpicture}
\end{frame}

@@

{ % all template changes are local to this group.
    \setbeamertemplate{navigation symbols}{}
    \begin{frame}<article:0>[plain]
        \begin{tikzpicture}[remember picture,overlay]
            \node[at=(current page.center)] {
                \includegraphics[keepaspectratio, height=\paperheight]{future.jpg}
            };
        \end{tikzpicture}
     \end{frame}
}

@@

\begin{frame}{Algebraic Solutions to Parameterised Problems (2/2)}
  With the help of a computer algebra system, we can then:
  \begin{itemize}
    \item determine how \alert{$P$} scales with the number of users,
    \item find combinations of domain sizes that keep \alert{$P$} below some
          threshold,
    \item find ranges of weights that keep \alert{$P$} sufficiently small across
          a range of domain size values.
  \end{itemize}
  \pause
  \vfill
  \centering
  \begin{tikzpicture}
    \node[draw] (a) {Reasoning with \alert{functions}};
    \node[right=0.5cm of a] (b) {\Huge\textcolor{color2}{$>$}};
    \node[draw,right=0.5cm of b] (c) {Reasoning with \alert{numbers}};
  \end{tikzpicture}
\end{frame}
